```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tips at Restaurants

In this homework we will use the `tips` data set. This data set is part of the `reshape2` package. You can load the data set by executing the command:
```{r}
data(tips, package="reshape2")
```
If you do not have available the package `reshape2`, issue `install.packages('reshape2')` to install it. The information contained in the data is collected by one waiter, who recorded over the course of a season information about each tip he received working in one restaurant. See `?tips` for a description of all of the variables. 

**Submission instructions**: Create a folder named `ds202_hw3`, and name the RMarkDown file `hw3.Rmd`  which should include your solutions under this folder. For submission, create a GitHub repository named `ds202_hw3` under your GitHub account, and push both `hw3.Rmd` and the knitted `hw3.html` before the deadline. I will assume you use the same GitHub username as for your HW2 submission. The latest version of your homework appearing on GitHub before the deadline is graded. *It is crucial to follow the exact folder structure and file names*, so that your homework can be reproduced and graded by an automated script. This homework is **due on Feb 19** before class.

1.  How many parties did the waiter serve? Store the number in `numParty` and print. 

```{r}
numParty <- nrow(tips)
numParty
```

2. What are the types of variables in the data frame `tips`? Include the code and also explain verbally.

```{r}
str(tips)
```

total_bill and tip are type "num" ||
sex, smoker, day, and time are type "Factor" ||
size is type "size"

3. Create a vector named `day` containing the day of visits in `tips`. The factor levels should be ordered from Thursday to Sunday. Print the variable.

```{r}
day <- levels(factor(tips$day, levels = c("Thur", "Fri", "Sat", "Sun")))
day
```

4. Create a data frame named `female5` containing the meal paid by a female payer in a party with size greater than or euqal to 5. Print the data frame.

```{r}
female5 <- tips[tips$sex == "Female" & tips$size >= 5,]
female5
```

5. How does the tipping amount (`tip`) depend on the overall bill (`total_bill`)? Use the `ggplot2` package to make a chart. Describe the relationship in words.  

```{r}
library(ggplot2)
ggplot(data = tips, aes(x = total_bill, y = tip)) +
  geom_point() + geom_smooth(method='lm') + geom_jitter()
```

The relation ship is positive and linear and has a considerable amount of variability.

6. Describe at least two types of anomalies in the previous plot. What do they mean?

One 'anomoly' is the tip of over $5 while the total bill was around $7.50, which is somewhere around a 66% tip. Very generous I would say.
Additionally, there is a very small tip with a bill near $32.5. Not so generous.

Another anomoly is the straight horizontal lines at around $2 and $3 for tips respectively, despite the bill amount changing. I am not sure why this would be.

7. Introduce a variable `tiprate` into the data set which stands for the rate of tips. What is the average rate for tips? Explain verbally. 

```{r}
tips$tiprate = tips$tip / tips$total_bill

mean(tips$tiprate)
```

The mean tip rate is .1608 or 16.08%. This means that on average, for every dollar that a bill increases, the tip will increase by around $0.16.

8. Make a boxplot of the tip rate by time. The x-axis should be ordered by lunch and then dinner. Use `ggplot2` to make a chart. Verbally explain the chart.

```{r}
tips$time <- factor(tips$time, levels = c("Lunch", "Dinner"))
ggplot(tips, aes(y = tiprate)) + geom_boxplot() + facet_wrap(~time)
```

The distribution of tip rate for both times looks similar, with the center as previously discussed, although the dinner tip rates have a slightly larger range as well as some outliers on the high side. Note that the lunch tip rates have no outliers.

9. How does smoking behavior and gender of the person who pays impact the relationship between tip and total bill? Find a visualization that incorporates all four variables. Interpret the result.

```{r}
ggplot(tips, aes(x = total_bill, y = tip)) + geom_point() + geom_smooth() + facet_grid(sex ~ smoker)
```

Non-smokers tend to have a more predictibly linear relatoinship between total bill and tip. Smokers have a less predictible tip pattern. Males appear to have more tips in the higher range than women. Smoking women have tips that taper off at the higher range, while male smokers a dip in tips near Q1 and Q3 of their total bill as well as a drasticincrease at the max.

10. Use ggplot2 to find a graphical summary of the relationship between day of the week and gender of the person paying the bill. What can you say about this relationship?

```{r}
tips$day <- factor(tips$day, levels = c("Thur", "Fri", "Sat", "Sun"))
ggplot(tips, aes(x = day, y = sex, size)) + geom_count()
```

Little visits by all sexes on fridays. Males have more visits than females on weekends, and females have a lower amount of visits on Sundays. Thursdays have a significant amount of traffic, split roughly equally among gender.

Note: your submission is supposed to be fully reproducible, i.e. the TA and I will 'knit' your submission in RStudio. 
